---
title: "mousedatacleaning"
output: html_document
date: "2025-01-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
getwd()
```
```{r packages + lilbararies}
# Install pacman 
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

# load packages
pacman::p_load(readxl, openxlsx, future, furrr, dplyr, tidyr)
```

# loading in the data and splitting into relevant chunks 
# we are working with a k-limbic file, which has some metadata at the beginning, trial data and some coded metadata. we will just use the trial
# data here, which is stored in references list. 
# adittionally, every file is one day, which contains multiple mice trials for that day
# in this code, i pretend all the days are one long trial for each mouse
# I also transform the punishments to be useful to my purposes, but feel free to change that 
# keep in mind the deck numbers have been switched on the mice, reference the excel document given by the Finnish researchers, 
# the code fixes this, but be aware of this when trying to debug.
# have fun!

```{r run this instead - load in these files into two lists instead}


library(readr)

# path
# folder_path <-  "/work/mousedata"

folder_path <-"/work/desccionmaking/mousedata"

# find all files
files <- list.files(folder_path, pattern = "\\.xlsx\\.rds$", full.names = TRUE)

# Separate files into references and datasets (named when I was confused, datasets are codes where all the activity is stored)
references_files <- files[grep("references_", files)]
datasets_files <- files[grep("datasets_", files)]

# Extract day
extract_pattern <- function(file_name) {
  pattern <- sub(".*_(\\d+_\\d+).*", "\\1", file_name)
  return(pattern)
}

# Sort files based on day
references_files <- references_files[order(sapply(references_files, extract_pattern))]
datasets_files <- datasets_files[order(sapply(datasets_files, extract_pattern))]

# Load the files into lists
references <- lapply(references_files, readRDS)
datasets <- lapply(datasets_files, readRDS)

# Names each index in the list based on the day of recording 
names(references) <- basename(references_files)
names(datasets) <- basename(datasets_files)

```

# finds id and cuts off so we only have choices 
```{r finds subject id}
library(data.table)

# store Subject Id 
subject_id_list <- list()

# Loop over the list of references

for (ref_name in names(references)) {
  day_references <- references[[ref_name]]
  
  # Loop over each experiment within the day
  for (j in seq_along(day_references)) {
    reference <- as.data.table(day_references[[j]])
    
    # Print the current reference
    #print(j)
    
    # Find the row where any column contains "Subject Id"
    subject_id_row <- which(reference == "Subject Id", arr.ind = TRUE)[, 1]
    
    # If the 'Subject Id' is found, extract the associated value from the next column
    if (length(subject_id_row) > 0) {
      subject_id_value <- reference[subject_id_row, which(reference[subject_id_row, ] == "Subject Id") + 1, with = FALSE]
      #print(subject_id_value)  # Preview the extracted value
      
      # Store the Subject Id value in the list with the index as the key
      subject_id_list[[ref_name]] <- c(subject_id_list[[ref_name]], subject_id_value)
    } else {
      print("NA")
    }
    
    # Save the updated reference back to the list
    day_references[[j]] <- reference
  }
  
  # saves the updated day references back to the main list
  references[[ref_name]] <- day_references
}

# Outputs the list of references and Subject Ids
#print(references)
#print(subject_id_list)
```

```{r cutting off the dataframes when we reach the Ref  in STARTDATA col since we got id}
# Loops through each index in references
for (j in 1:length(references)) {
  # loops through each dataframe in the current index
  for (k in 1:length(references[[j]])) {
    # extracts the current dataframe from the index
    df1 <- references[[j]][[k]]
    
    # Checks if 'STARTDATA' column  (col 1) exists in the dataframe
    if ("STARTDATA" %in% colnames(df1)) {
      # Find the position of the first occurrence of 'Ref' in STARTDATA
      trial_start_row <- which(df1$STARTDATA == "Ref")[1]
      
      # checks if there is a 'Ref'
      if (!is.na(trial_start_row)) {
        # snips everything off starting from ref
        df1 <- df1[trial_start_row:nrow(df1), ]
        
        # updates df 
        references[[j]][[k]] <- df1
      } else {
        cat("no ref in df ", k, " in sublist ", j, ". Skipping.\n")
      }
    } else {
      cat("miv no startdata col", k, " in sublist ", j, ". Skipping.\n")
    }
  }
}



```
```{r taking out the time collumn from the original data}
library(data.table)

extract_column_data <- function(nested_list) {
  lapply(nested_list, function(inner_list) {
    if (!is.list(inner_list)) return(NULL) 
    
    lapply(inner_list, function(dt) {
      if (!inherits(dt, "data.table")) return(NULL)  #there are a lot of checks here because it kept throwing errors, idk why but this helps
      
      # finds column "...5"
      col_name <- grep("\\.\\.\\.5$", names(dt), value = TRUE) 
      
      if (length(col_name) == 1) {
        return(dt[[col_name]][-1])  # excludes the first row
      } else {
        return(NULL)  # if its not there, skip
      }
    })
  })
}

time_list <- extract_column_data(references)



```

```{r getting the time per turn}
time_list <- lapply(time_list, function(inner_list) {
  lapply(inner_list, function(vec) {
    if (!is.null(vec) && length(vec) > 1) {
      return(diff(as.numeric(vec)))  # this is quite nice bc the time is in one col.. sometimes we are lucky
    } else {
      return(NULL)  # when one trial ends, the next starts immediately after, which allows us to do this
    }
  })
})

```


```{r deck choices and removing last row if any NA}
# Moves everything into a new list where we record the deck selections
results_list <- setNames(vector("list", length(references)), names(references))

# go through all ref
for (ref_name in names(references)) {
  # get names
  ref_dataframes <- references[[ref_name]]
  
  # list
  results_list[[ref_name]] <- vector("list", length(ref_dataframes))
  
  # go through all the df found
  for (df_index in seq_along(ref_dataframes)) {
    # take one
    df <- ref_dataframes[[df_index]]
    
    # for some reason nessecary, also doing this in data table instead of data frame for speed 
    df <- as.data.table(df)
    
    # remove last row if na, sometimes this one is recorded, but doesnt contain any data, so out it goes 
    if (any(is.na(df[nrow(df), ]))) {
      df <- df[-.N]  
    }

    # making a new place to store deck choice
    results_df <- data.table(deck = rep(0, nrow(df)))

    # finds the deck choice per trial
    for (row_index in seq(2, nrow(df))) {
      if (!is.na(df[row_index, "...32", with = FALSE][[1]]) && df[row_index, "...32", with = FALSE][[1]] != 0) {
        results_df$deck[row_index] <- 1
      } else if (!is.na(df[row_index, "...33", with = FALSE][[1]]) && df[row_index, "...33", with = FALSE][[1]] != 0) {
        results_df$deck[row_index] <- 2
      } else if (!is.na(df[row_index, "...34", with = FALSE][[1]]) && df[row_index, "...34", with = FALSE][[1]] != 0) {
        results_df$deck[row_index] <- 3
      } else if (!is.na(df[row_index, "...35", with = FALSE][[1]]) && df[row_index, "...35", with = FALSE][[1]] != 0) {
        results_df$deck[row_index] <- 4
      } else {
        results_df$deck[row_index] <- 0  # Default to 0 if no nonzero value is found
      }
    }
    
    # removes the first row of the deck collumn, since it has words instead of data (forgot to cut this off earlier)
    results_df <- results_df[-1, ]  
    
    # Store
    results_list[[ref_name]][[df_index]] <- results_df
    
    # Save 
    references[[ref_name]][[df_index]] <- df
  }
}

```


```{r mergning the two lists} 
# adds time, keeps date as name for index
results_list <- setNames(lapply(names(results_list), function(i) {
  setNames(lapply(seq_along(results_list[[i]]), function(j) {
    results_list[[i]][[j]][, Time := time_list[[i]][[j]]]
  }), names(results_list[[i]]))  
}), names(results_list)) 
```



# return to here to add id when we are done selecting the selected decks 

```{r looping over multiple dataframes and adding ID}

# Check if lengths of results_list and subject_id_list match
if (length(results_list) != length(subject_id_list)) {
  stop("nope")
}

#adds id to all of the mice
for (ref_index in seq_along(results_list)) {
  # Iterate through each data.table within the current reference
  for (df_index in seq_along(results_list[[ref_index]])) {
    # Get the current data.table from results_list
    dt <- results_list[[ref_index]][[df_index]]
    
    #check
    setDT(dt)
    
    # gets id 
    subject_id <- subject_id_list[[ref_index]][[df_index]]
    
    # add it onto our deck choice dfs
    dt[, ID := subject_id]
    
    # i like it to be the first col
    setcolorder(dt, c("ID", setdiff(names(dt), "ID")))
    
    # save
    results_list[[ref_index]][[df_index]] <- dt
  }
}

```


# adding if they got a reward or not 

```{r recording reward and adding to results}


for (j in 1:length(references)) {
  for (k in 1:length(references[[j]])) {
    df1 <- references[[j]][[k]]
    
    df2 <- results_list[[j]][[k]]
    
    setDT(df2)
  
    # prepare reward 
    if (!"reward" %in% names(df2)) {
      df2[, reward := 0]
    }
    
    # Loop through each row starting from the 2nd row in df1
    for (i in 2:nrow(df1)) {  # Start loop from row 2 (second row) in df1
      # checks the columns 45 to 48 and converts  to numeric, NA=0
      cols <- as.numeric(c(df1[i, "...45", with = FALSE], df1[i, "...46", with = FALSE], 
                           df1[i, "...47", with = FALSE], df1[i, "...48", with = FALSE]))
      cols[is.na(cols)] <- 0  # Convert NA to 0
      
      # Findsw non-zero values 
      non_zero_values <- cols[cols %in% c(1, 2, 3, 4)]
      
      # If any add to df 
      if (length(non_zero_values) > 0) {
        df2[i - 1, reward := non_zero_values[1]]  
      } else {
        df2[i - 1, reward := 0]  # 0 if none found
      }
    }
    
    # save
    results_list[[j]][[k]] <- df2
  }
}


```

```{r changing the deck selections given the knowledge about participant id}


for (i in seq_along(results_list)) {
 
  for (j in seq_along(results_list[[i]])) {
  
    dt <- results_list[[i]][[j]]
    
    setDT(dt)
    dt[, ID := as.numeric(ID)]
    
    # is ID is odd or even
    if (dt$ID[1] %% 2 == 1) {
      # odd ids
      dt[, deck := case_when(
        deck == 1 ~ 1,  # 1 becomes 1
        deck == 2 ~ 4,  # 2 becomes 4
        deck == 3 ~ 2,  # 3 becomes 2
        deck == 4 ~ 3   # 4 becomes 3
      )]
    } else {
      # even ids
      dt[, deck := case_when(
        deck == 1 ~ 4,  # 1 becomes 4
        deck == 2 ~ 1,  # 2 becomes 1
        deck == 3 ~ 3,  # 3 becomes 3
        deck == 4 ~ 2   # 4 becomes 2
      )]
    }
    
    # save
    results_list[[i]][[j]] <- dt
  }
}


```

```{r removing rows wher the mice did not choose anything}

for (i in seq_along(results_list)) {
  
  for (j in seq_along(results_list[[i]])) {
   
    dt <- results_list[[i]][[j]]
    
    setDT(dt)
    
    # removes when mice blanked a run - this might be nice to keep if doing more complex stuff
    dt <- dt[!is.na(deck)]
    
    # save
    results_list[[i]][[j]] <- dt
  }
}


```


```{r adding in the trial number}

# Iterate over each list in results_list
for (i in seq_along(results_list)) {
 
  for (j in seq_along(results_list[[i]])) {
    
    setDT(results_list[[i]][[j]])
    
    # adds a trial, aka choice number
    results_list[[i]][[j]][, trial := .I]
  }
}

```


```{r find the dates from the index}
dates_index <- sapply(names(results_list), function(name) {
  # get date of run
  sub(".*_(\\d{1,2}_\\d{1,2})\\.xlsx\\.rds", "\\1", name)
})
```

```{r get date}
# add the date column
for (i in seq_along(results_list)) {

  current_date <- dates_index[i]
  
  for (j in seq_along(results_list[[i]])) {
    results_list[[i]][[j]]$date <- current_date
  }
}
```

```{r getting reward per second 🔥}
total_reward_all <- 0
total_time_all <- 0

for (i in seq_along(results_list)) {
  
  for (dt in results_list[[i]]) {
    
    dt$Time <- as.numeric(as.character(dt$Time))
    dt$reward <- as.numeric(as.character(dt$reward))
    
    # NA is 0
    dt$Time[is.na(dt$Time)] <- 0
    dt$reward[is.na(dt$reward)] <- 0
    
    # only get reward rows
    dt <- dt[dt$reward %in% c(1, 2, 3, 4), ]
    
    # removes time 0 
    dt <- dt[dt$Time > 0, ]
    
    # adds all reward and time up
    total_reward_all <- total_reward_all + sum(dt$reward, na.rm = TRUE)
    total_time_all <- total_time_all + sum(dt$Time, na.rm = TRUE) / 100  # Convert centiseconds to seconds
  }
}

# average reward per second
overall_avg_reward_per_second <- total_reward_all / total_time_all

```
```{r adding a punishment collumn}
# different punishments on the decks, so need to multiply to get the right ones
punishment_multipliers <- c("1" = 5, "2" = 10, "3" = 30, "4" = 40)

# same same, looking back some of these loops could have been merged rip
for (i in seq_along(results_list)) {
  
  for (j in seq_along(results_list[[i]])) {
    dt <- results_list[[i]][[j]]
    
    dt$deck <- suppressWarnings(as.numeric(dt$deck))
    dt$reward <- suppressWarnings(as.numeric(dt$reward))
    
    dt$punishments <- 0
    
    #finds punishment rows
    zero_reward_rows <- dt$reward == 0 & !is.na(dt$deck)
    
    # get the punishment based on the deck choices
    dt$punishments[zero_reward_rows] <- -overall_avg_reward_per_second * punishment_multipliers[as.character(dt$deck[zero_reward_rows])]
    
    # save
    results_list[[i]][[j]] <- dt
  }
}

```

```{r collect all datatables again}

all_data_tables <- list()

#this collapses the nested structure to begin to merge the mice, so each mouse has all their trials together

for (i in seq_along(results_list)) {
  
  if (is.list(results_list[[i]])) {
    all_data_tables <- c(all_data_tables, results_list[[i]])
  } else {
    all_data_tables <- c(all_data_tables, list(results_list[[i]]))
  }
}

```


```{r new list, same mice on same index}

same_mouse <- list()

#finds id 
first_row_ids <- sapply(all_data_tables, function(df) df$ID[1])

unique_ids <- unique(first_row_ids)

# group by id
for (id in unique_ids) {

  matching_tables <- which(first_row_ids == id)
  
  # save 
  same_mouse[[as.character(id)]] <- all_data_tables[matching_tables]
}

```

```{r merging the same mouse into one dataframe}
merged_list <- lapply(same_mouse, bind_rows)

```

```{r mergning dataframes into one}
individual_mouse_dataframes <- lapply(merged_list, as.data.frame)

```

```{r in between to get date and month into two cols}

# split the date column
split_date_column <- function(df) {
  df <- df %>%
    mutate(
      day = as.numeric(sub("_.*", "", df[["date"]])),  #day
      month = as.numeric(sub(".*_", "", df[["date"]])) # month
    )
  return(df)
}

individual_mouse_dataframes <- lapply(individual_mouse_dataframes, split_date_column)


```

```{r sorting to get one long trial count instead of per sessison}

create_choice_column <- function(df, df_name) {
  
  #one df is empty so just skip, 
  if (is.null(df) || nrow(df) == 0) {
    warning(paste("Warning: Dataframe", df_name, "is NULL or empty. Skipping."))
    return(df) 
  }
  
  # Sorting by month, day, and trial
  df <- df %>%
    arrange(month, day, trial) %>%
    mutate(choice = row_number())  # generates row numbers in order of day
  
  return(df)
}


individual_mouse_dataframes <- lapply(names(individual_mouse_dataframes), function(df_name) {
 
  if (!is.null(individual_mouse_dataframes[[df_name]])) {
    individual_mouse_dataframes[[df_name]] <- create_choice_column(individual_mouse_dataframes[[df_name]], df_name)
  }
  return(individual_mouse_dataframes[[df_name]])
})

```

```{r renaming cols to match with the one used in class script}
# clean and merge 
process_and_merge <- function(df_list) {
  # remove empty df
  df_list <- Filter(function(df) is.data.frame(df) && nrow(df) > 0, df_list)
  
  if (length(df_list) == 0) {
    print("NOPE")
    return(NULL)
  }

  merged_df <- do.call(rbind, lapply(df_list, function(df) {
    
    df <- df[, c("ID", "deck", "reward", "punishments", "choice"), drop = FALSE]
    
    #rename cols
    colnames(df) <- c("subjID", "deck", "gain", "loss", "trial")
    
    return(df)
  }))
  
  return(merged_df)
}


# merge dfs
IGT_sober_mouse <- process_and_merge(individual_mouse_dataframes)


```


```{r cutting down so equal numnber of trials per id}
# find minimum number of trials per id
# Ensure all IDs have exactly 100 trials
IGT_sober_mouse <- IGT_sober_mouse %>%
  group_by(subjID) %>%
  slice_head(n = 100) %>%  #keep first 100 rows per ID
  ungroup()


```

```{r save as txt}

save_merged_txt <- function(df, filename) {
    write.table(df, file = filename, sep = "\t", row.names = FALSE, quote = TRUE)
}


# Save as one TXT file
save_merged_txt(IGT_sober_mouse, "IGT_sober_mouse.txt")
```


